---
title: "edgeR with different models"
author: "Nisha paudel"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: simplex
    highlight: pygments
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Experiments with combinations of multiple factors

<p>Instructions for the code are from https://f1000research.com/articles/5-1438 and edgeR bioconductor manual and chapter 3.3 page 36 of edgeR manual https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf </p>

<p>Based on edgeR quasilikelihood pipeline, NB model with F tests -because of strict error control rate than traditional edgeR and DESeq2 -limma also offer excellent error control and are recommended for large datasets due to its speed</p>

<p>The data is from a paper Generation of a microglial developmental index in mice and in humans reveals a sex difference in maturation and immune reactivity Published in Glia in 2017.</p>

<p>Here is analysis for P60 sal, E18, P4, P14 for both male and female. We are using design <- model.matrix(~ 0 + Group), where group is interation Sex * Age,  this design is easy to implement and recommended by authors
is equivalent to design <- model.matrix(~ 0 + Sex + Sex:Age, data=targets)/(~ 0+ Sex +Sex:Age) <b> this is called Nested factor Design </b>
page 34 


## load libraries and data
```{r}
library(edgeR)
library(limma)
count_data1 <- read.csv('/Users/nishapaudel/Documents/CASP/Data/Counts.csv', sep =',', header=T, row.names = 1)
sampleinfo <- read.csv("/Users/nishapaudel/Documents/CASP/Data/NEW_design.txt", sep = ',', header = TRUE, row.names = 1, stringsAsFactors = TRUE)
sampleinfo
dim(count_data1)
```

## Grouping for interation
```{r}
group <- factor(paste0(sampleinfo$Sex, ".", sampleinfo$Age))
y <- DGEList(count_data1, group=group)
```

## Filter
```{r}
keep <- filterByExpr(y)
summary(keep)
```

## Normalize
```{r}
y <- normLibSizes(y) # Default normalization is TMM
y$samples
```
## MDS
```{r}
plotMDS(y, col= as.numeric(group))
```



## Model Design
```{r}
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)
design
```
## Dispersion estimation
```{r}

y <- estimateDisp(y, design, robust=TRUE)
y$common.dispersion
plotBCV(y)


```
## Fitting to GLM
```{r}
fit <- glmQLFit(y, design, robust=TRUE)
head(fit$coefficients)
```

```{r}
plotQLDisp(fit)
```


## ANOVA-like testing -Sex wise DEGs at various age groups

```{r}
con <- makeContrasts(F18vsM18 = F.E18 - M.E18, FP4vsMP4  = F.P4 - M.P4,FP14vsMP14  = F.P14 - M.P14, FP60_SvsMP60_S = F.P60_S - M.P60_S, levels=design)

```



```{r}
anov <- glmQLFTest(fit, contrast=con)
topTags(anov)
```


```{r}
con <- makeContrasts(F18vsM18 = F.E18 - M.E18, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_E18_sex <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_E18_sex <-rownames(subset(sig,logFC<0.4))
```


```{r}
con <- makeContrasts( FP4vsMP4  = F.P4 - M.P4, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_P4_sex <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_P4_sex <-rownames(subset(sig,logFC<0.4))
```


```{r}
con <- makeContrasts(FP14vsMP14  = F.P14 - M.P14, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_P14_sex <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_P14_sex <-rownames(subset(sig,logFC<0.4))
```


```{r}
con <- makeContrasts(FP60_SvsMP60_S = F.P60_S - M.P60_S, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_P60_sex <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_P60_sex <-rownames(subset(sig,logFC<0.4))
```

## VennDiagram Sex wise up regulated DEGs
Kdm5d and Dcpp1 are the single unique genes on E18 and P60

Overlapping genes are Tsix and Xist which are present in Female

```{r}
library(VennDiagram)
# Combine the lists into a named list for the Venn diagram
venn_data <- list("E18"=dge_edgr_up_E18_sex,"P 4"=dge_edgr_up_P4_sex,"P 14"=dge_edgr_up_P14_sex,"P 60"=dge_edgr_up_P60_sex )

# Create the Venn diagram with custom colors
vennColors <- c("lightpink", "cyan1", "violet","lightgreen" )

# Set up the Venn diagram configuration
venn_config <- venn.diagram(
  x = venn_data,
  category.names = names(venn_data),
  fill = vennColors,
  filename = NULL, # Set filename to NULL to display the plot directly in R session
  imagetype = "png", # You can change this to other formats like "pdf", "jpeg", etc.
  col = "black", # Set background color to transparent for better display in R session
  cex = 2.7,
  cex.label = 7,
  lwd = 2 
)

# Display the Venn diagram
grid.newpage()
grid.draw(venn_config)
```



## VennDiagram Sex wise down regulated DEGs
Overlapping genes are Uty, Kdm5d, Ddx3y and Eif2s3y.
```{r}
library(VennDiagram)
# Combine the lists into a named list for the Venn diagram
venn_data <- list("E18"=dge_edgr_down_E18_sex,"P 4"=dge_edgr_down_P4_sex,"P 14"=dge_edgr_down_P14_sex,"P 60"=dge_edgr_down_P60_sex )

# Create the Venn diagram with custom colors
vennColors <- c("lightpink", "cyan1", "violet","lightgreen" )

# Set up the Venn diagram configuration
venn_config <- venn.diagram(
  x = venn_data,
  category.names = names(venn_data),
  fill = vennColors,
  filename = NULL, # Set filename to NULL to display the plot directly in R session
  imagetype = "png", # You can change this to other formats like "pdf", "jpeg", etc.
  col = "black", # Set background color to transparent for better display in R session
  cex = 2.7,
  cex.label = 7,
  lwd = 2 
)

# Display the Venn diagram
grid.newpage()
grid.draw(venn_config)
```

## ANOVA-like testing -Age wise DEGs at various age groups
## For Female 

```{r}
con <- makeContrasts(F18vsF4 = F.E18 - F.P4, FP4vsFP14  = F.P4 - F.P14,FP14vsFP60_S  = F.P14 - F.P60_S, levels=design)
```

```{r}
anov <- glmQLFTest(fit, contrast=con)
topTags(anov)
```
```{r}
con <- makeContrasts(F18vsF4 = F.E18 - F.P4, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
plotMD(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_F_18_VS_4 <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_F_18_VS_4 <-rownames(subset(sig,logFC<0.4))
```


```{r}
con <- makeContrasts( FP4vsFP14  = F.P4 - F.P14, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_F_14VS4 <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_F_14VS4 <-rownames(subset(sig,logFC<0.4))
```


```{r}
con <- makeContrasts(FP14vsFP60_S  = F.P14 - F.P60_S, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_F_14VS60 <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_F_14VS60 <-rownames(subset(sig,logFC<0.4))
```

## VennDiagram Female age wise up regulated DEGs
```{r}
# Combine the lists into a named list for the Venn diagram
venn_data <- list("F.E18 VS F.P4"=dge_edgr_up_F_18_VS_4,"F.P4 VS F.P14"=dge_edgr_up_F_14VS4,"F.P14 VS F.P60"=dge_edgr_up_F_14VS60 )

# Create the Venn diagram with custom colors
vennColors <- c("lightpink", "cyan1", "violet" )

# Set up the Venn diagram configuration
venn_config <- venn.diagram(
  x = venn_data,
  category.names = names(venn_data),
  fill = vennColors,
  filename = NULL, # Set filename to NULL to display the plot directly in R session
  imagetype = "png", # You can change this to other formats like "pdf", "jpeg", etc.
  col = "black", # Set background color to transparent for better display in R session
  cex = 2.7,
  cex.label = 7,
  lwd = 2 
)

# Display the Venn diagram
grid.newpage()
grid.draw(venn_config)
```
## VennDiagram Female age wise down regulated DEGs 
```{r}

# Combine the lists into a named list for the Venn diagram
venn_data <- list("F.E18 VS F.P4"=dge_edgr_down_F_18_VS_4,"F.P4 VS F.P14"=dge_edgr_down_F_14VS4,"F.P14 VS F.P60"=dge_edgr_down_F_14VS60 )

# Create the Venn diagram with custom colors
vennColors <- c("lightpink", "cyan1", "violet" )

# Set up the Venn diagram configuration
venn_config <- venn.diagram(
  x = venn_data,
  category.names = names(venn_data),
  fill = vennColors,
  filename = NULL, # Set filename to NULL to display the plot directly in R session
  imagetype = "png", # You can change this to other formats like "pdf", "jpeg", etc.
  col = "black", # Set background color to transparent for better display in R session
  cex = 2.7,
  cex.label = 7,
  lwd = 2 
)

# Display the Venn diagram
grid.newpage()
grid.draw(venn_config)
```




###  age wise for male


```{r}
con <- makeContrasts(M18vsM4 = M.E18 - M.P4, MP4vsMP14  = M.P4 - M.P14,MP14vsMP60_S  = M.P14 - M.P60_S, levels=design)
```

```{r}
anov <- glmQLFTest(fit, contrast=con)
topTags(anov)
```

```{r}
con <- makeContrasts(M18vsM4 = M.E18 - M.P4, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
plotMD(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_M_18_VS_4 <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_M_18_VS_4 <-rownames(subset(sig,logFC<0.4))
```

```{r}
con <- makeContrasts( MP4vsMP14  = M.P4 - M.P14, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_M_14VS4 <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_M_14VS4 <-rownames(subset(sig,logFC<0.4))
```



```{r}
con <- makeContrasts(MP14vsMP60_S  = M.P14 - M.P60_S, levels=design)
qlf <- glmQLFTest(fit, contrast= con )
topTags(qlf)
summary(decideTests(qlf))
table(p.adjust(qlf$table$PValue, method = "BH") < 0.05)
dge<-topTags(qlf, n=Inf)
dge_edgR <-dge
sig <- subset(dge_edgR$table,FDR < 0.05)
dge_edgr_up_M_14VS60 <-rownames(subset(sig,logFC>0.4))
dge_edgr_down_M_14VS60 <-rownames(subset(sig,logFC<0.4))
```
### VennDiagram Male age wise up regulated DEGs
```{r}
# Combine the lists into a named list for the Venn diagram
venn_data <- list("M.E18 VS M.P4"=dge_edgr_up_M_18_VS_4,"M.P4 VS M.P14"=dge_edgr_up_M_14VS4,"M.P14 VS M.P60"=dge_edgr_up_M_14VS60 )

# Create the Venn diagram with custom colors
vennColors <- c("purple", "grey", "lightgreen" )

# Set up the Venn diagram configuration
venn_config <- venn.diagram(
  x = venn_data,
  category.names = names(venn_data),
  fill = vennColors,
  filename = NULL, # Set filename to NULL to display the plot directly in R session
  imagetype = "png", # You can change this to other formats like "pdf", "jpeg", etc.
  col = "black", # Set background color to transparent for better display in R session
  cex = 2.7,
  cex.label = 7,
  lwd = 2 
)

# Display the Venn diagram
grid.newpage()
grid.draw(venn_config)
```
### VennDiagram Male age wise down regulated DEGs
```{r}
# Combine the lists into a named list for the Venn diagram
venn_data <- list("M.E18 VS M.P4"=dge_edgr_down_M_18_VS_4,"M.P4 VS M.P14"=dge_edgr_down_M_14VS4,"M.P14 VS M.P60"=dge_edgr_down_M_14VS60 )

# Create the Venn diagram with custom colors
vennColors <- c("purple", "grey", "lightgreen" )

# Set up the Venn diagram configuration
venn_config <- venn.diagram(
  x = venn_data,
  category.names = names(venn_data),
  fill = vennColors,
  filename = NULL, # Set filename to NULL to display the plot directly in R session
  imagetype = "png", # You can change this to other formats like "pdf", "jpeg", etc.
  col = "black", # Set background color to transparent for better display in R session
  cex = 2.7,
  cex.label = 7,
  lwd = 2 
)

# Display the Venn diagram
grid.newpage()
grid.draw(venn_config)
```
## For reproducibility

```{r}
sessionInfo()
```

```{r}
getwd()
```

